Space Matters: extending sensitivity analysis to initial spatial conditions in geosimulation models

### 2018-09-19 Response_to_Reviewers.tex

	% 6. Line 232: comparing phase diagrams is related to the topic of comparing patterns. It would make sense to mention the methods that are usually used to compare patterns
	\item \comment[JR]{biblio a rajouter}

	% 7. Line 238: Please explain why this particular index was chosen and discuss its properties. Connect the discussion to other indices like L1, and Linf norms, etc. Would we get qualitatively different results from different indices? 
	\item \comment[JR]{on avait teste avec le Monge ca donnait a peut pres pareil ; donc les normes alpha surement encore plus. je regarde.}

	% 8. Regarding the spatial generator: the authors should explain why this generator is relevant to these two models? 
	\item This spatial generator is relevant to both model as it produces density grids at a ``metropolitan scale'' (see detailed discussion added in Appendix and the original paper from which we used the generator), which is the scale at which both model were initially intended to be. We made this aspect more explicit.

	\item In the case of Schelling, this is the scale at which most empirical segregation indices are computed. In the case of Sugarscape, it corresponds to the whole city if the model is a metaphor for city resources (Batty....), or to a generic landscape where a resource is grown otherwise. In both cases, our point is that there exist many different patterns of density distribution in resource location and urban density, and that not acknowledging this diversity leads to potentially false results. Furthermore, in urban models, we argue that the hypothesis of isotropic density is potentially the most harmful one (e.g. the least realistic, although this is the case of most Schelling models).
	%\comment[CC]{Ref du papier de batty à ajouter (je ne l'ai pas trouvée}[(c'est le Batty, 2005)]

@@

	% 9. The authors should illustrate the contribution of each parameter of the generator to the generated initial pattern using a figure. Details of the generators should be included in an appendix
	\item The behavior of the spatial generator was described with more details in Appendix.

	% 10. Explain how the density grids are used in the Schelling model, do you allow more than one agent to be in the same cell? 
	\item Yes, we allow more agents to be in the same cell. The potential density is defined by the density grid generated, and its population with the same number of agents in all grids. This distribution is the one at initialisation. If the potential density is not reached, more agents can move in the cell, otherwise it is deemed full and unavailable for further moves. The satisfaction and segregation indices are computed with regard to the people in the cell and the people present in neighbouring cells.
	%\comment[JR]{il a mal lu ou on est pas assez explicites, a detailler}
	%\comment[CC]{ça va comme réponse? }[(JR) parfait]
	%\comment[CC]{argumentaire à ajouter dans le papier}[(cc)done]

	% 11. Line 297: I am not sure why empirical density destitutions of cities are important in the context of this study. 
	%\item cf infra. 
	\item Density distributions of empirical cities are important because we want to test models with realistic ranges of initial patterns of density distribution. Therefore we cannot limit ourselves to a single 2-peak mountain in the classical Sugarscape model and an isotropic square city in Schelling. We chose to use actual distribution of European cities to constraint our density generation.
%\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]

@@

%This is an interesting article that I read with true interest. The paper fits the profile of this journal and can reach a wider audience.

%I also have some concerns:

\begin{itemize}
	% In the abstract the authors claim that initial spatial conditions are often taken for granted. I do not agree to this statement. Initial spatial condition can have a large impact on the results of a model, however I do not think that the initial spatial conditions are often taken for granted.
	\item We forcefully agree with the referee in that initial spatial conditions can have a large impact on the results of the model. Maybe not all papers take them for granted, but what we meant is that compared to other parameters, initial spatial conditions are much less often explored using sensitivity analysis. Actually, we know only a few papers which have explored the impact of initial spatial conditions as a systematic criterion in a sensitivity analysis of model results. We clarified this point in the introduction.
    %\comment[JR]{il s'agirait de nuancer ici}
	%\comment[CC]{argumentaire à ajouter dans le papier}[(cc) done]

	% Page 2 line 88: the authors state that “Even in applied cases when GIS geometries of a particular city are used, the spatial distribution of agents is still approximated by a constant density”, I do not think this is true. Many models generate a heterogeneous synthetic population that is based on local data (replicating the actual population as closely as possible) and in far more detail then what is done in this paper. I am thinking of work in the line of disease diffusion, crime etc.
	\item \comment[JR]{il a un point car ce sont des classes de modèles qu'on a pas trop cité ? par contre la sensibilite a ces distribs precises n'est pas testée, ou pas systeématiquement : il faut qu'on clarifie ou reformule}
    \comment[MLT]{En fait, c'est là aussi une question de reformulation. Je crois qu'il faudrait répondre quelque chose comme : "We acknowledge the fact that many ABM applications use "real" data and spatial features for their simulations. However, as explained by (Thomas et al. 2017) in the case of LUTI models for instance, most simulations do not evaluate the impact of city delineations nor MAUP on the output of the models, which constitutes a potential limit in the interpretation of results.}{(JR) ca me va comme reformulation}

@@

	% The most interesting part of the research to me, is not the way the density layers are generated but the use of the phase diagrams, defined as a vector of final aggregated model outputs. As these are vectors, they are not so different from time series and many techniques exist to compare vectors in this domain. I wonder if the Euclidean distance is the best measure to determine the difference between the phase vectors of different spatial inputs.
	% -> 
	\item The comment on other possible techniques to compare phase diagrams echoes one other reviewer comment, and we indeed expanded the discussion on this point, showing first that other distance do not change much qualitative results and with more literature for possible extensions.
	%\comment[JR]{analyse à ajouter}

	% Last paragraph of page 5, you do not state explicitly if genetic algorithms were use within the OpenMOLE platform.
	\item The functionalities of OpenMOLE we used were detailed; we did not use genetic algorithms but sampling methods and high performance computing environments.

	% In section 3.1, page 8, line 404-405 the authors mention that the distance of the phase diagrams was measured with respect to the reference phase diagram computed on the default initial spatial conditions. Actually this came as a surprise to me. I had the impression that all phase diagrams would be evaluated against all other diagrams. Perhaps this should be explained earlier in the paper.
	\item \comment[JR]{ce qu'il suggère c'est un indice intégré sur l'ensemble des phases diagramme, donc un ``meta-meta-indice'' en quelque sorte, nous on reste à un niveau de meta seulement car on regarde juste pour chaque s'il bouge par rapport à un point de reference. En termes thematique c'est plus coherent pour nous, lui doit avoir en tete des indices de sensibilite globaux type Saltelli. $\rightarrow$ a voir comment justifier ca bien}{(JR) inclure en mise en contexte ce lien avec les Saltelli et indexes globaux}

@@

	% I appreciate the fact that the authors in section 4 clearly indicate the limitations of their work (and I share the limitations listed under “comparing phase diagrams”). Personally I do not think that releasing the density grids is an important opportunity. As mentioned before I think that the core of the work is in the comparison of the phase vectors. I am not so sure the second opportunity listed in 4.2 is very valid. I do not think that the CEUS journal targets social sciences exactly.
	\item Thanks for the remarks on discussion, in particular on releasing density grids. We have tried to shorten the considerations which you have found not too useful and will let the readers decide if the grids could be of any use to them.
\end{itemize}

Concerning the comments made by the third referee:

\begin{itemize}
	% I found the research question posed by the author(s) interesting and I agree with the author(s) that many modelers don’t consider how spatial arrangements impact the outcomes of the model and tend to focus only on the parameters. The paper makes good use of the complexity and spatial literature in defining the problem and demonstrates how urban form impacts the results. However, while the paper is innovative, I found it really difficult to interpret and follow the results. I would suggest to the author to rewrite the results and methodology so that they flow better and also provide more context with respect to how to interpret the results.
	\item We have tried to make the methodology and result section flow better in the corresponding sections.
    \comment[CC]{à faire en derniere relecture}

	% On a side note, one of your key words you use “validation” but in what sense?  You don’t really come back to this in the paper, also you write about “the validity of generative models is uncertain until their results are proven robust and representative of ’real-world’ conditions.” In the abstract but this is it.  What do you consider real world conditions to be? You seem to allude that you need to have the spatial generator to generate realistic (styles) city shapes for testing on the Schelling model but I am not convinced.  After the abstract all notion of validation is missing.
	\item We have agreed with the reviewer on ``validation'' and removed the concept of validation altogether, replacing it with model evaluation or sensitivity analysis.
    %\comment[JR]{enlever le concept de validation}

@@

	% Geosimulation is introduced but not defined or referenced.  This should be clarified as the term has several meanings etc.
	\item For the definition of geosimulation, we meant the one coined by (Benenson and Torrens, 2004), this was made explicit and the reference was added.

	% In places the English along with the overall argument could be better articulated. For example, on page 1 you write: “We think simulation can become a very good tool to achieve this,”  Could this be rewritten or better argued. You are not the only one to believe this but you don’t show this in your paper.
	\item We have worked towards improving the English along with the overall argument. 
      \comment[CC]{ok pour cette phrase + derniere relecture a faire}

	% Page 4: you write “Generally, sensitivity analyses have focused either on the spatial context (extent of the environment, shape of the zones if applicable - squares, hexagons, etc., objects - links, rasters, etc.); either on the spatial encoding of the heterogeneity of space (algorithms of disaggregation, interpolation, etc.), but rarely on both at the same time” This makes little sense. What it should be is “Generally, sensitivity analyses have focused either on the spatial context or the heterogeneity of space” or something like that. Basically you need the “or”.
	\item Thank you for the suggestion on page 4. We have modified the paragraph accordingly.

	% You write “To our knowledge there exist no single well established method to compare phase diagrams in the agent-based modelling and geosimulation literature.” As this is new you may want to discuss how say its done in other fields?  
	\item 
    \comment[JR]{cf remarques other reviewers sur ce point}[(MLT) Sa remarque fait sens, mais j'avoue ne pas être trop au point sur cette partie de la littérature... vous avez des références?][(JR)concernant la spatial sensitivity : justement c'est pas trop fait non plus dans les autres champs ! je parlais avec un ecologue, leur espace est jamais varié autre que donnees reelles ; j'ai l'impressions que geosciences, hydrologie etc idem. Pour comparer phase diagrams j'ai pas en tete de domaine ou ca se ferait (faute d'avoir cherché, justement quand on cherchait une methode) - le plus proche c'est les patterns comparison etc. comme l'on suggere les reviewers.]

@@

\end{itemize}

Concerning the comments made by the fourth referee:

\begin{itemize}
	% In sensitivity analysis of spatial models, stochastic effects and small parameter changes are normally applied. What about initial conditions? The authors test the effects of different arrangements (e.g. monocentric vs. polycentric density patterns) on agent-based Schelling segregation models and Sugarscape's unequal societies model. Overall, I find this to be a very good paper. Very well written in general. Good review of the problem, clearly outlining (with references) the need for focus on initial conditions in geosimulation. The chief factors (data uncertainty, nature of spatial units (e.g. shape) and spatial heterogeneity) on initial spatial conditions are covered: -how many of these do you test? I see heterogeneity addressed, but what about shape and uncertainty? You touch upon shape when talking about NetLogo's use of raster as opposed to vector in the Discussion - could you expand on this?
	\item We have not addressed the shape of cells and the uncertainty of data in this paper, only the diversity of shapes of urban morphologies (e.g. heterogeneity).

	% Rigorous experimental method described. In the matching of European metropolitan areas to your generated outputs, I wonder what you are missing by not using cities from other continents. i.e. by widening the cities available to your study it may be possible to even better discover real-world examples of your archtypes (compact, polycentric, discontinuous).
	\item To the best of our knowledge, there does not exist a study of urban morphologies similar to the one we used but at a global scale (this should be linked to the absence of a raster of population with a resolution fine enough). Working on other regional systems of cities such as the US could be possible but remains out of the scope of this paper. We added this interesting suggestion to the discussion.
	%\comment[JR]{to the best of my knowledge there is no full-world morphological classification of cities (and surely because global 1km density dataset is not robust enough for that - already with the European grid it is not perfect)}[(MLT) Peut-être qu'on peut intégrer sa remarque dans la partie discussion? Je ne connais pas non plus de classification morphologique des villes à l'échelle du monde. Les cas européens sont privilégiés du fait de la disponibilité de données comme Corine Land Cover et Urban Atlas (pour l'usage du sol) ou Geostat (grilles de densité de pop de résolution 1km), mais ça doit aussi exister pour les US au moins? Il y a aussi le projet gridded population of the world de l'université de Columbia mais la résolution n'est pas assez fine (2.5 arcminute) pour une analyse de la morpho urbaine.]

@@

	% Why do you not have the same amount of repetitions for your Sugarscape (50 repetitions) and Schelling (100 repetitions) model runs?
	\item \comment[JR]{no justification for the difference in number of runs (just experimental setup contingence\ldots ) $\rightarrow$ either new runs or redo computations with 50 for Schelling}\comment[JR]{ca va etre chaud patate pour les new computations ; je vais regarder les ratio de sharpe, avec un peu de chance il y a besoin de plus de repets pour schelling...}

	% You say your models are open source - do you give details of that source?  
	\item Concerning the source code of the models, an open public repository was put into place and the link provided in the paper (to an anonymized version to ensure blind review). Similarly, the large simulation datasets used for results presented in the paper are openly available on a dataverse repository (provisorily anonymized also).
	%\comment[JR]{TODO dataverse. / statement open code and data}

### 2018-09-19 paper.tex
.

\usepackage{listings}
\usepackage{mdframed}

@@

\todo{add code and data availability statement here}

@@

\paragraph{Comparing phase diagrams} Comparing phase diagrams is as we saw not so straightforward, and further developments of our method imply testing alternative methods for this particular point. For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \deleted{\emph{meta phase transitions}} phase transitions in the space of \replaced{generator}{meta} parameters. The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\citep{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools. Map comparison methods, popular in environmental sciences, provide numeral tools to compare two dimensional fields~\citep{visser2006map}. To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\citep{10.1371/journal.pone.0178165}. The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.

\added{\cite{kuhnert2005comparing} map comparison techniques}

\added{\cite{kruskal1983overview} parallel with time-series comparison techniques (see reviewer comment - however more constrained as two dimensional indexing)}

@@

Although the applications we present here are limited by the simplicity of the models, we think that the method could (and should) be applied to larger models including domain mechanisms and more empirical initialisation data, for example synthetic populations. The sensitivity analysis to initial spatial conditions could then be either a replication on the spatial allocation of the synthetic population, or a series of spatial permutations of the empirical spatial inputs.
We want to foster this extension of our work by releasing the density grids also generated, as well as the generating work-flow and the model implementation. They are available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}. Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents.

@@

@@

\added{We give in Fig.~\ref{fig:stylized-generator} plots roughly summarizing the behavior of the density grid generator, according to parameters $(\alpha,\beta)$.}

\todo{bla bla on generator scale / real configs in the initial paper}

%%%%%%%%%%%%%%
\begin{figure*}
	\includegraphics[width=0.49\linewidth]{figures/alpha_beta_examples.png}
    \includegraphics[width=0.49\linewidth]{figures/alpha_beta_schemas.png}
    \caption{\added{\textbf{Stylized behavior of the density grid generator.}}\label{fig:stylized-generator}}
\end{figure*}
%%%%%%%%%%%%%%

@@

\added{The Fig.~\ref{fig:schelling-distance-meta} gives the phase diagrams distances as a function of generator parameters and morphological components, similarly to the Sugarscape model in main text. In absolute, this version of the Schelling model seems less sensitive to density grids than the sugarscape model, as we do not obtain a high range of values here.}

%%%%%%%%%%%%%
\begin{figure*}[h!]
\centering
\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams_red}
\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace_red}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} Each point corresponds to a spatial configuration and colour gives the relative distance to one of the phase diagrams. We show them in the \replaced{generator }{meta-}parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
\end{figure*}
%%%%%%%%%%%%%

@@

\added{We give below the pseudo-code for the implementations we used of both models.}

\begin{figure*}
%\begin{frame}
% Minimal netlogo code for the sugarscape model
\begin{lstlisting}
	to main-loop
    	regrow-patches
        update-agetns
    end

    to regrow-patches
    	ask patches [set psugar min (list max-psugar (psugar + 1))]
    end

    to update-agents
    	ask turtles [
    		let move-candidates (patch-set patch-here (patches at-points vision-points) with [not any? turtles-here])
  let possible-winners move-candidates with-max [psugar]
  if any? possible-winners [
    ;; if there are any such patches move to one of the patches that is closest
    move-to min-one-of possible-winners [distance myself]
  ]

            ; 
            set sugar (sugar - metabolism + psugar)
  			set psugar 0

    		set age (age + 1)
    		if sugar <= 0 or age > max-age [
      			hatch 1 [ new-turtle ]
      			die
    		]
  		]
    end

\end{lstlisting}
%\end{frame}
\end{figure*}

%%%%%%%%%%%%%%%
\begin{figure*}
	\begin{lstlisting}

    \end{lstlisting}
\end{figure*}
%%%%%%%%%%%%%%%

>>>>

\usepackage{listings}
\lstset{escapeinside={(*@}{@*)}}

\usepackage{mdframed}

@@

\added{The implementations of the models were done in NetLogo. We modified the sugarscape version with wealth of NetLogo model library (to be able to explore it intensively) and we implemented from scratch the Schelling model. Both pseudo-codes are available in Appendix~\ref{app:code}, and source code for models, grid classification and simulation results analysis is available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}. Density grids are also available at this address. Simulation data are available for reproducibility on the dataverse repository at \url{https://doi.org/10.7910/DVN/A0N5GV}.}

@@

\paragraph{Comparing phase diagrams} Comparing phase diagrams is as we saw not so straightforward, and further developments of our method imply testing alternative methods for this particular point. For example in the case of the Schelling model, an anisotropic spatial segregation index (giving the number of clusters found and in which region in the parameter spaces they are roughly situated) would differentiate strong \deleted{\emph{meta phase transitions}} phase transitions in the space of \replaced{generator}{meta} parameters. The use of metrics comparing spatial distributions, such as the Earth Movers Distance which is used for example in Computer Vision to compare probability distributions~\citep{rubner2000earth}, or the comparison of aggregated transition matrices of the dynamic associated to the potential described by each distribution, would also be potential tools. Map comparison methods, popular in environmental sciences, provide numeral tools to compare two dimensional fields~\citep{visser2006map,kuhnert2005comparing}. To compare a spatial field evolving in time, elaborated methods such as Empirical Orthogonal Functions that isolates temporal from spatial variations, would be applicable in our case by taking time as a parameter dimension, but these have been shown to perform similarly to direct visual inspection when averaged over a crowdsourcing~\citep{10.1371/journal.pone.0178165}. \added{The transfer of methods used to compare sequences \citep{kruskal1983overview} or time-series \citep{liao2005clustering} is a possible way to develop measures between phase diagrams. The higher dimension of the phase diagram we study must however be considered with caution when transferring method, in a way analog to the application of global sensitivity indices to spatial data \citep{lilburne2009sensitivity}.} The investigation of diverse approaches to systematically quantify differences between phase diagrams is an important potential development of our method.

@@

Although the applications we present here are limited by the simplicity of the models, we think that the method could (and should) be applied to larger models including domain mechanisms and more empirical initialisation data, for example synthetic populations. The sensitivity analysis to initial spatial conditions could then be either a replication on the spatial allocation of the synthetic population, or a series of spatial permutations of the empirical spatial inputs.
We want to foster this extension of our work by releasing the density grids also generated, as well as the generating work-flow and the model implementation. They are available on the open repository of the project at \url{https://github.com/AnonymousAuthor2/SpaceMatters}. \added{Future work could be done to compare these or generate grids with a larger morphological span, covering other typical urban forms that can be found in the world.}

Another way to go would be to implement additional generators, such as social networks \citep{alizadeh2016generating} with localised agents, \added{transportation networks generators \citep{raimbault2018multi}, or coupled road network and population raster generators \citep{raimbault2018urban}.}

@@

\newpage

@@

\added{We give in Fig.~\ref{fig:stylized-generator} plots roughly summarizing the behavior of the density grid generator, according to parameters $(\alpha,\beta)$. To put it in a simple way, high values of $\alpha$ give highly hierarchical configurations, and diminishing $\beta$ increase the number of centers. Low values of $\alpha$ give diffuse patterns, with however clear centers for a high diffusion. We do not discuss it the role of other parameters, but according to \cite{raimbault2018calibration}, diffusion steps give smoother forms, and the rate between total population and the population increment at each step (which is equivalent to the total number of steps) is crucial to select non-stationary distributions that are closer to real configurations. \cite{raimbault2018calibration} also shows the existence of non-linear behaviors in some regions of the parameter space, so the description we gave here shall not be interpreted as a linear link between generator parameters and the morphological properties of the generated grids.}

%%%%%%%%%%%%%%
\begin{figure*}
	\includegraphics[width=0.49\linewidth]{figures/alpha_beta_examples.png}
    \includegraphics[width=0.47\linewidth]{figures/alpha_beta_schemas.png}
    \caption{\added{\textbf{Stylized behavior of the density grid generator.} (Left) Examples of grids with their position in the $(\alpha,\beta)$ parameter space; (Right) Stylized interpretation of the forms obtained as a function of $\alpha$ and $\beta$.}\label{fig:stylized-generator}}
\end{figure*}
%%%%%%%%%%%%%%

@@

\added{The Fig.~\ref{fig:schelling-distance-meta} gives the phase diagrams distances as a function of generator parameters and morphological components, similarly to the Sugarscape model in main text. In absolute, this version of the Schelling model seems less sensitive to density grids than the sugarscape model, as we do not obtain a high range of values here. We however obtain measures ranging from 0.3 to 1, what is however characteristic of a significant sensitivity to space.}

%%%%%%%%%%%%%
\begin{figure*}
\centering
\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_metaparams_red}
\includegraphics[width=0.48\textwidth]{figures/schelling-relativedistance_morphspace_red}
\caption{\textbf{Relative distances of phase diagrams to the reference across grids for the Schelling model.} Each point corresponds to a spatial configuration and colour gives the relative distance to one of the phase diagrams. We show them in the \replaced{generator }{meta-}parameter space (Left) and in the reduced morphological space (Right).\label{fig:schelling-distance-meta}}
\end{figure*}
%%%%%%%%%%%%%

\section{\added{Comparison of phase diagram with other distances}}

\added{We describe here the tests done with other distances to compare phase diagrams.}

@@

\added{We give below the pseudo-code for the implementations we used of both sugarscape and schelling models. We recall that source code is openly available at} \url{https://github.com/AnonymousAuthor2/SpaceMatters}.

%%%%%%%%%%%%%%%
\begin{figure*}
	\caption{\added{\textbf{Pseudo-code (NetLogo style) for the Schelling model used here.}}}
	\begin{mdframed}
	\begin{lstlisting}
	while (*@\textit{not stopping-condition}@*) [
    	ask turtles [
    		set unsatisfied? false
    		let neighboragents turtles in-radius neighborhood-radius
    		let wantedcolor color
    		let stranger-rate count neighboragents with [color = wantedcolor] / count neighboragents
    		if stranger-rate < similar-wanted [
    set unsatisfied? true random-move
    ]
  ]
    \end{lstlisting}
    \end{mdframed}
\end{figure*}
%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%
\begin{figure*}
% Minimal netlogo code for the sugarscape model
\caption{\added{\textbf{Pseudo-code (NetLogo style) for the sugarscape model.} We give a pseudo-code very close to the NetLogo language, at the exception of italic expressions which are not put here in NetLogo to ease understandability.}}
\begin{mdframed}
\begin{lstlisting}
while [time < final-time] [
	; regrow the patch resources
    ask patches [set psugar min (list max-psugar (psugar + 1))]

    ; make agents exploit resources
    ask turtles [
    	; candidate patches to which the turtle can move
    	let move-candidates (*@\textit{patches in vision range with no turtles}@*)

    	; select the patch with maximal resource
 		let possible-winners move-candidates with-max [psugar]
  		if any? possible-winners [
  			move-to one-of (*@\textit{possible-winners at minimal distance}@*)
  		]

        ; eat the sugar on the patch
        set sugar (sugar - metabolism + psugar)
  		set psugar 0

        ; age and die if necessary
    	set age (age + 1)
    	if sugar <= 0 or age > max-age [
    		; hatch an other turtle before dying
      		hatch 1 [ new-turtle ]
      		die
    	]
  	]
]
\end{lstlisting}
\end{mdframed}
\end{figure*}
%%%%%%%%%%%%%%%
